Evaluating infectious disease forecasts with allocation scoring rules

### 2023-08-30 alloscore_manuscript/allocation.tex
easy edits in response to Aaron's comments

Infectious disease forecasting models have emerged as important tools in public health. Predictions of disease dynamics have been used to inform decision-making about a wide variety of measures to reduce disease spread and/or mitigate the severity of disease outcomes. For example, estimates of expected onset of flu season have been used to aid national vaccination strategies \citep{igboh2023timing}, and forecasts of Ebola dynamics have been used to allocate surveillance resources \citep{meltzer2014estimating, rainisch2015regional}. \cite{bertsimas2021predictionsCOVID} developed tools to inform decision making from infectious disease forecasts, which have been used to inform allocation of limited medical supplies such as ventilators, ICU capacity planning, and vaccine distribution strategy. Models developed by \cite{fox_real-time_2022} have been used to inform resource and care site planning, as well as community guidelines for masking, traveling, dining and shopping (University of Texas, 2022) \nocite{utnews2022}. In April of 2022, the Centers for Disease Control and Prevention (CDC) announced the launch of the Center for Forecasting and Outbreak Analytics (CFA) to translate disease forecasts into decision-making (CDC, 2022)\nocite{cdc2022cfa}, indicating that this has been identified as an important direction at the highest levels of government public health response. The value of infectious disease forecasts has typically been measured by how closely they predict disease outcomes such as cases, hospitalizations or deaths using, for example, root mean square error (RMSE) \citep{papastefanopoulos2020covid} or weighted interval score (WIS) \citep{bracher2021evaluating}, however, recently authors have been calling for evaluating forecasts through their impact on policy \citep{marshall2023predictions}.

In decision-making settings where it is possible to quantify the utility or loss associated with a particular action, standard tools of decision theory provide a procedure for developing forecast scoring rules that measure the value of forecasts through the quality of the decisions that they lead to. We give an overview of these procedures in Section \ref{sec:methods.decisiontheory}. There is a large history of literature applying these ideas to obtain measures of the value of forecasts that are tied to a decision-making context, primarily in fields such as economics and finance, supply chain management, and meteorology. We review this work only briefly here, and we refer the reader to \cite{yardley2021utility_cost_forecasts} for a general overview, and to \cite{pesaran2002decision_based_eval} and \elr{TODO: identify relevant review or book style summary for meteorology} for discussions focused on applications to economics and meteorology, respectively. In finance, the value of forecasts can often be measured by the profits generated by trading decisions informed by the forecasts, perhaps adjusted for risk levels \cite[e.g.,][]{leitch1991economicForecastEval, cenesizoglu2012returnPredictionEconValue}. In applications to supply chain management and meteorology, the value of forecasts has typically been operationalized by considering the costs associated with decisions regarding the amount of inventory to hold or the level of protection against the impacts of extreme weather events to enact {Peter Catt (2007), Fotios Petropoulos and colleagues (2019), Nada Sanders and Gregory Graman (2009), T.N. Palmer (2002), Florian Pappenberger and colleagues (2015)}. For example, in supply chain management these decisions may incur costs related to holding inventory, labor, or providing poor service, while in meteorology we may need to balance the costs of implementing protective measures with the costs of potentially preventable weather damages. In this framework, a forecast has value if it leads to decisions with low total costs. In all of these fields, analyses have consistently found that common measures of statistical forecast accuracy do not necessarily correspond directly to measures of the value of forecasts as an input to decision-making \cite[e.g.,][]{leitch1991economicForecastEval, cenesizoglu2012returnPredictionEconValue}. % Consistently, but not uniformly

% Other important early contributions include Nelson (1972) and White (1966), who are amongst the first to consider the problem of forecast evaluation from the perspective of decision theory and dynamic programming. Nelson, in particular, articulates the concept of the “value of a (probability) fore- cast” to an individual decision-maker as the expected value of his/her payoff using the probability forecasts relative to the payoff expected if unconditional probability estimates are used.

@@

Briefly, the allocation score of a forecast is the avoidable unmet need that results from using that forecast to set resource allocations by minimizing expected unmet need.
For example, suppose that a decision-maker is provided with forecasts of the level of need for medical resources in each of several states or hospital systems.
If there is a limited amount of the medical resource that is available to distribute, a decision-maker could choose an allocation of that resource across locations that minimizes the expected unmet need according to that forecast.
As measured by the allocation score, one forecast is better than another if it would lead decision-makers to an allocation that results in less unmet need.
%By ``unnecessary'' we mean the unmet need that could have been avoided by an oracle that knows exactly how much need will occur in each location and divides the amount $K$ so that nothing is wasted in one location while it could be put to use in another.
If the amount of resources that is available to distribute is less than the actual need, some amount of unmet need is unavoidable.
The allocation score for a forecast does not include the unmet need that was unavoidable given the resource constraint, and so it measures only the amount of unmet need that could have been prevented by using a different allocation of available resources than that suggested by the forecast.
We elaborate on these ideas in Section~\ref{sec:methods}.

We present an illustrative analysis using the allocation score to evaluate forecasts of hospital admissions in the US leading up to the Omicron wave in winter 2022.
This analysis is ``synthetic'' in that it does not correspond to an actual analysis that supported decision-making in real-time.
However, the framework described in this paper corresponds to real-world decisions that must be made by public health administrators around the globe, and could be adapted in the future for such real-time situations.
For example, forecasts for districts in Sierra Leone of bed demand to care for patients with Ebola was the subject of a real-time modeling study in late 2014 and early 2015 \citep{camacho2015-ebola-bed}.
And, in 2020, a model developed by an academic research group turned predictions of COVID-19 hospitalizations into estimates of ventilator usage and and shortages. This framework was used by the Hartford HealthCare system in Connecticut ``to align ventilator supply with projected demand at a time where the [COVID-19] pandemic was on the rise'' \citep{bertsimas2021predictionsCOVID}.
These examples illustrate the potential for forecasts to inform decisions about how to allocate limited supplies such as temporary hospital beds, ventilators, personal protective equipment, or other supplies that are known to be effective at reducing transmission or severity of disease.
However, we emphasize again that these studies did not take the step of evaluating forecasts based on the quality of the allocation decisions that they supported or could have been used to support.

@@

We begin with an informal description of the allocation score and some examples illustrating its key characteristics in section \ref{sec:methods.overview}. In section \ref{sec:methods.detailed} we develop the allocation score more carefully, building on decision-theoretic procedures for deriving proper scoring rules. We then comment on some connections between the allocation score that we propose and other common scores that can be derived from decision-theoretic foundations, such as the quantile score, WIS, and CRPS, in section \ref{sec:methods.related}.

@@

Suppose that a decision-maker is tasked with determining how much of a resource such as a medical supply should be allocated to each of $N$ locations, subject to the constraint that only $K$ units of the resource are available.
If the decision-maker is provided with a multivariate forecast $F$ that predicts resource need in each location, one option is to choose the resource allocation that minimizes the expected total unmet need according to the forecast.
We will give a more precise mathematical statement in section \ref{sec:methods.allocation}, but informally, the total expected unmet need according to the forecast is
\begin{align}
\sum_{i=1}^N E_{F_i}[\text{unmet need in location $i$}], \label{eqn:informal_objective}
\end{align}
where $F_i$ is the marginal forecast distribution of need in location $i$ and the unmet need in a particular location is the difference between resource need in that location and the number of resources that were allocated there.
This allocation problem has an intuitively appealing solution: allocate so that the probabilities of need exceeding allocation in various locations are as close to each other as possible.
This will lead to the allocations provided by $F$ being quantiles of the marginal distributions $F_i$ for some \emph{single} probability level $\tau$ that is shared in common for all locations.

@@

This illustrates a fundamental property of the allocation score: at its core, it measures whether the forecast accurately captures the relative magnitudes of resource need across different locations, which is precisely the information that is needed to allocate resources to those locations subject to a fixed resource constraint.
On the other hand, the allocation score is not directly sensitive to whether the forecasts in each location correctly capture the magnitude of resource need in each individual location.
%For a given probability level $\tau$, if the quantiles $Q_i(\tau)$ of the marginal forecasts are proportional to the observed needs $y_i$, then the allocation score is zero for the resource constraint level $K = \sum Q_i(\tau)$.
This stands in marked contrast to other common scoring methods for multivariate forecasts that aggregate univariate scores such as CRPS or WIS for the marginal forecasts where a 
misalignment in one coordinate is penalized regardless of alignments in other coordinates. Note that we do not claim that the allocation score is generically preferable to these other scores \textemdash rather, it provides a view of forecast performance that is specifically tuned to the context of decision-making about resource allocations.

@@

\subsection{A decision-theoretic development of the allocation score}
\label{sec:methods.detailed}

We give a high-level review of a general procedure for developing proper scoring rules that are tailored to specific decision-making tasks in section \ref{sec:methods.detailed.decisiontheory}, and then in section \ref{sec:methods.detailed.allocation} we apply that procedure to develop the allocation score based on the task of deciding on how to allocate a fixed supply of resources across multiple locations.

\subsubsection{The decision-theoretic setup for forecast evaluation}
\label{sec:methods.detailed.decisiontheory}

%In this section, we give an overview of the decision-theoretic setup for developing proper scoring rules that measure the value of a forecast as an input to decision making. We keep the discussion here at a somewhat informal level; we refer the reader to [some subset of Brehmer and Gneiting; Grünwald and Dawid; Dawid; Granger and Pesaran 2000; Granger and Machina 2006; Ehm et al. 2016] for more technically precise discussion.

In the framework of decision theory, a decision corresponds to the selection of an action $x$ from some set of possible actions $\mathcal{X}$. For example, $x$ may correspond to the level of investment in a measure designed to mitigate severe disease outcomes such as hospital beds, ventilators, medication, or medical staff, with $\mathcal{X}$ being the set of all possible levels of investment that we might select. The quality of a decision to take a particular action $x$ is measured in relation to an outcome $y$ that is unknown at the time the decision is made. For example, $y$ may correspond to the number of individuals who eventually become sick and would benefit from the mitigation measure, and informally, an action $x$ is successful to the extent that it meets the realized need. In the face of uncertainty, a decision-maker may use a forecast $F$ of the random variable $Y$ to help inform the selection of the action to take. We measure the value of a forecast as an input to this decision-making process by the quality of the decisions that it leads to.

We can formalize the preceding discussion with the following three-step procedure for developing scoring rules for probabilistic forecasts:
\begin{enumerate}
\item Specify a \emph{loss function} $s(x, y)$ that measures the loss associated with taking action $x$ when outcome $y$ eventually occurs.
\item Given a probabilistic forecast $F$, determine the \emph{Bayes act} $x^F$ that minimizes the expected loss under the distribution $F$.
\item The \emph{scoring rule} for $F$ calculates the score as the loss incurred when the Bayes act was used: $S(F, y) = s(x^F, y)$.
\end{enumerate}
%We use the letter $s$ for the loss function to align with the literature on evaluation of forecasts of continuous outcomes, in which context we can often identify the action $x$ with a functional (i.e., a numeric summary such as a mean or a quantile) of the forecast distribution $F$. In this context, $s$ may be used as a \emph{scoring function}. \elr{Consider moving preceding sentences to a footnote or just deleting them?}
This is a general procedure that may be applied in settings where it is possible to specify a quantitative loss function. Subject to certain technical conditions, scoring rules obtained from this procedure are proper. We refer the reader to [some subset of Brehmer and Gneiting; Grünwald and Dawid; Dawid; Granger and Pesaran 2000; Granger and Machina 2006; Ehm et al. 2016] for more technically precise discussion.

@@

% We review how the quantile score arises from a particular decision-making problem in section \ref{sec:methods.quantileloss.quantile}, and how CRPS can be obtained by integrating across values of the parameters of that decision-making problem, as well as the connection to WIS, in section \ref{sec:methods.quantileloss.crps}. These results have been thoroughly discussed in the literature [cite cite cite].

% \subsubsection{Decision-theoretic origins of the quantile score}
% \label{sec:methods.quantileloss.quantile}

% Suppose that a decision-maker is tasked with determining the quantity $x$ of a protective measure to procure; for example, $x$ might represent the number of hospital beds or amount of medicine to purchase. Additionally, suppose that each unit of this good has cost $C$ so that the total cost of procurement is $Cx$. The variable $y$ denotes the eventual realized need for this resource, e.g. the number of patients in need of a hospital bed or the amount of medication that is needed. We assume that each unit of unmet need incurs a loss denoted by $L$, so that if the selected procurement level $x$ is less than the realized need $y$, a loss of $L(x-y)$ results. At the time that a decision-maker determines the amount $x$ to procure, the demand $y$ is not yet known. We therefore define the random variable $Y$ that represents the as-yet-unknown level of demand. The forecast $F$ specifies a predictive distribution for $Y$. Here we identify $F$ with its cumulative distribution function (CDF), and $F^{-1}$ denotes the quantile function. With this formalization of the decision-making task, we can proceed to develop a proper scoring rule using the procedure outlined in section \ref{sec:methods.decisiontheory}.

@@

% We have set up the problem here in terms of the cost and loss parameters $C$ and $L$, which has the benefit of an intuitive connection to the decision-making context. However, to clarify the connection to the usual notation for the quantile loss, we can divide the loss function $s_Q$ by $L$ to obtain an expression in terms of only $\alpha$:
% \begin{align*}
% s_Q(x,y; \alpha) &= s_Q(x,y; C, L)/L \\
% &= (C/L)x + (x-y)_- \\
% &= (1 - \alpha)x + (x-y)_-.
% \end{align*}
% Because these loss functions are equal up to a constant of proportionality, the Bayes act is the same for both. The associated quantile scoring rule expressed in terms of $\alpha$ is
% \begin{equation}
% S_Q(F, y; \alpha) = (1 - \alpha)F^{-1}(\alpha) + \left(F^{-1}(\alpha) - y\right)_-. \label{eqn:S_Q_alpha}
% \end{equation}
% In either formulation, the key observation is that the Bayes act is the quantile of the forecast distribution $F$ at the probability level given by one minus the cost/loss ratio $C/L$.

@@

% The scoring rule $S_Q$ of Equation \eqref{eqn:S_Q_alpha} evaluates the forecast distribution $F$ only through its $\alpha$ quantile. While this is faithful to the context of the decision-making problem, it may not be satisfying as a measure of the quality of the full forecast distribution. For this purpose, one option is to integrate the quantile scoring rule across different values of the probability level $\alpha$, weighting the probability levels according to a specified distribution $p$. This yields a weighted CRPS:
% \begin{align*}
% S_{CRPS}(F, y; p) &= \int S_Q(F, y; \alpha) p(\alpha) \, d\alpha.
% \end{align*}
% This weighted form of CRPS has appeared in the literature before, e.g. see \cite{gneiting2011weightedScoringRules}. The usual CRPS results from taking $p$ corresponding to a $\text{Uniform}(0,1)$ distribution, equally weighting all probability levels.

% We emphasize that because $\alpha = 1 - C/L$, the distribution $p(\alpha)$ can be interpreted as expressing incomplete knowledge about the cost/loss ratio in the decision-making problem. The equal weighting used by the ordinary CRPS may be appropriate in the absence of any knowledge about the context in which forecasts will be used to support decision-making, but may be inappropriate if more information is known about the cost/loss ratio for a specific decision-making task.

@@

%As a concrete example, we take the resource to be a good such as ventilators or oxygen supply. An administrator is tasked with determining where to send these resources so as to meet demand among hospital patients in different facilities or states. In contrast to the decision-making problem in the previous section, the administrator is not able to control the total amount of supply; rather, their task is to determine how to allocate the fixed supply to different locations. Following the structure of the discussion for the quantile score and CRPS, we first discuss the set up for a single value of the resource constraint $K$, which plays a role that is analogous to the cost/loss ratio for the quantile score, and then we discuss the possibility of integrating (or averaging) across values of $K$.

@@

In the decision-making setting that we consider, an action $\mathbf{x} = (x_1, \ldots, x_N)$ is a vector specifying the amount that is allocated to each of $N$ locations. We require that each $x_i$ is non-negative and that the total allocation across all locations equals the amount of available resources, $K$: $\sum_{i=1}^N x_i = K$. The set $\mathcal{X}$ consists of all possible allocations that satisfy these constraints. The eventually realized resource need in each location is denoted by $\mathbf{y} = (y_1, \ldots, y_N)$. These levels of need are not known at the time of decision making, so we define the random vector $Y = (Y_1, \ldots, Y_N)$ where $Y_i$ represents the as-yet-unknown level of resource need in location $i$. Forecasts of need in each location are collected in $F = (F_1, \ldots, F_N)$. We assume that resource need is non-negative and the forecasts reflect that, i.e. the support of each $F_i$ is a subset of $\mathbb{R}^+$. Finally, we assume that each unit of unmet need incurs a loss denoted by $L$, so that if the selected resource level $x_i$ in location $i$ is less than the realized need $y_i$, a loss of $L(y_i - x_i)$ results.

@@

One interpretation of this result is that the Bayes act sets the allocation in each location $i$ to a quantile of the forecast distribution $F_i$ for that location. The quantile is at a probability level $\alpha = 1 - \lambda^{\star}/L$ that is the same for all locations, and is chosen such that the constraint is satisfied. According to this interpretation, we divide the available resources among the locations in such a way that the forecasted probability that need will exceed demand is the same in all locations.
An alternative interpretation comes from noting that for each location $i$, $\frac{\partial}{\partial x_i} \sbar_A^F(\mathbf{x}) \vert_{\mathbf{x}=\mathbf{x}^F} = \lambda^{\star}$ (see the supplement for a proof).
%$\frac{\partial}{\partial x_i} \sbar_F(\mathbf{x}) = L[F_i(x_i) - 1]$ (see the supplement for a derivation), and evaluating at the Bayes allocation we arrive at
%\begin{align}
%\frac{\partial}{\partial x_i} \sbar_F(\mathbf{x}) \vert_{\mathbf{x}=\mathbf{x}^F} &= L[F_i(x^F_i) - 1] = L[F_i(F_i^{-1}(1 - \lambda^{\star}/L)) - 1] = \lambda^{\star}.
%\end{align}
In words, at the allocation given by the Bayes act, the rate of change of the expected score as a function of the amount allocated to location $i$ is given by $\lambda^{\star}$.
This derivative is the same for all locations, so the optimal allocation divides the available resources across all locations in such a way that according to $F$, the expected benefit of 1 additional unit of resources is the same in all locations.

@@

Case study heading into the Omicron wave. Some more detailed discussion of implications of bad forecasts for specific decision-making purposes -- take a "deep dive" into one or two example states like FL.

>>>>

Infectious disease forecasting models have emerged as important tools in public health. Predictions of disease dynamics have been used to inform decision making about a wide variety of measures to reduce disease spread and/or mitigate the severity of disease outcomes. For example, estimates of expected onset of flu season have been used to aid national vaccination strategies \citep{igboh2023timing}, and forecasts of Ebola dynamics have been used to allocate surveillance resources \citep{meltzer2014estimating, rainisch2015regional}. \cite{bertsimas2021predictionsCOVID} developed tools to inform decision making from infectious disease forecasts, which have been used to inform allocation of limited medical supplies such as ventilators, ICU capacity planning, and vaccine distribution strategy. Models developed by \cite{fox_real-time_2022} have been used to inform resource and care site planning, as well as community guidelines for masking, traveling, dining and shopping (University of Texas, 2022) \nocite{utnews2022}. In April of 2022, the Centers for Disease Control and Prevention (CDC) announced the launch of the Center for Forecasting and Outbreak Analytics (CFA) to translate disease forecasts into decision making (CDC, 2022)\nocite{cdc2022cfa}, indicating that this has been identified as an important direction at the highest levels of government public health response. The value of infectious disease forecasts has typically been measured by how closely they predict disease outcomes such as cases, hospitalizations or deaths using, for example, root mean square error (RMSE) \citep{papastefanopoulos2020covid} or weighted interval score (WIS) \citep{bracher2021evaluating}, however, recently authors have been calling for evaluating forecasts through their impact on policy \citep{marshall2023predictions}.

In decision making settings where it is possible to quantify the utility or loss associated with a particular action, standard tools of decision theory provide a procedure for developing forecast scoring rules that measure the value of forecasts through the quality of the decisions that they lead to. We give an overview of these procedures in Section \ref{sec:methods.decisiontheory}. There is a large history of literature applying these ideas to obtain measures of the value of forecasts that are tied to a decision making context, primarily in fields such as economics and finance, supply chain management, and meteorology. We review this work only briefly here, and we refer the reader to \cite{yardley2021utility_cost_forecasts} for a general overview, and to \cite{pesaran2002decision_based_eval} and \elr{TODO: identify relevant review or book style summary for meteorology} for discussions focused on applications to economics and meteorology, respectively. In finance, the value of forecasts can often be measured by the profits generated by trading decisions informed by the forecasts, perhaps adjusted for risk levels \cite[e.g.,][]{leitch1991economicForecastEval, cenesizoglu2012returnPredictionEconValue}. In applications to supply chain management and meteorology, the value of forecasts has typically been operationalized by considering the costs associated with decisions regarding the amount of inventory to hold or the level of protection against the impacts of extreme weather events to enact {Peter Catt (2007), Fotios Petropoulos and colleagues (2019), Nada Sanders and Gregory Graman (2009), T.N. Palmer (2002), Florian Pappenberger and colleagues (2015)}. For example, in supply chain management these decisions may incur costs related to holding inventory, labor, or providing poor service, while in meteorology we may need to balance the costs of implementing protective measures with the costs of potentially preventable weather damages. In this framework, a forecast has value if it leads to decisions with low total costs. In all of these fields, analyses have consistently found that common measures of statistical forecast accuracy do not necessarily correspond directly to measures of the value of forecasts as an input to decision making \cite[e.g.,][]{leitch1991economicForecastEval, cenesizoglu2012returnPredictionEconValue}. % Consistently, but not uniformly

% Other important early contributions include Nelson (1972) and White (1966), who are amongst the first to consider the problem of forecast evaluation from the perspective of decision theory and dynamic programming. Nelson, in particular, articulates the concept of the “value of a (probability) fore- cast” to an individual decision maker as the expected value of his/her payoff using the probability forecasts relative to the payoff expected if unconditional probability estimates are used.

@@

Briefly, the allocation score of a forecast is the avoidable unmet need that results from using that forecast to set resource allocations by minimizing expected unmet need.
For example, suppose that a decision maker is provided with forecasts of the level of need for medical resources in each of several states or hospital systems.
If there is a limited amount of the medical resource that is available to distribute, a decision maker could choose an allocation of that resource across locations that minimizes the expected unmet need according to that forecast.
As measured by the allocation score, one forecast is better than another if it would lead decision makers to an allocation that results in less unmet need.
%By ``unnecessary'' we mean the unmet need that could have been avoided by an oracle that knows exactly how much need will occur in each location and divides the amount $K$ so that nothing is wasted in one location while it could be put to use in another.
If the amount of resources that is available to distribute is less than the actual need, some amount of unmet need is unavoidable.
The allocation score for a forecast does not include the unmet need that was unavoidable given the resource constraint, and so it measures only the amount of unmet need that could have been prevented by using a different allocation of available resources than that suggested by the forecast.
We elaborate on these ideas in Section~\ref{sec:methods}.

We present an illustrative analysis using the allocation score to evaluate forecasts of hospital admissions in the US leading up to the Omicron wave in winter 2022.
This analysis is ``synthetic'' in that it does not correspond to an actual analysis that supported decision making in real-time.
However, the framework described in this paper corresponds to real-world decisions that must be made by public health administrators around the globe, and could be adapted in the future for such real-time situations.
For example, forecasts for districts in Sierra Leone of bed demand to care for patients with Ebola was the subject of a real-time modeling study in late 2014 and early 2015 \citep{camacho2015-ebola-bed}.
And, in 2020, a model developed by an academic research group turned predictions of COVID-19 hospitalizations into estimates of ventilator usage and and shortages. This framework was used by the Hartford HealthCare system in Connecticut ``to align ventilator supply with projected demand at a time where the [COVID-19] pandemic was on the rise'' \citep{bertsimas2021predictionsCOVID}.
These examples illustrate the potential for forecasts to inform decisions about how to allocate limited supplies such as temporary hospital beds, ventilators, personal protective equipment, or other supplies that are known to be effective at reducing transmission or severity of disease.
However, we emphasize again that these studies did not take the step of evaluating forecasts based on the quality of the allocation decisions that they supported or could have been used to support.

@@

We begin with an informal description of the allocation score and some examples illustrating its key characteristics in section \ref{sec:methods.overview}. In section \ref{sec:methods.detailed} we develop the allocation score more carefully, building on decision theoretic procedures for deriving proper scoring rules. We then comment on some connections between the allocation score that we propose and other common scores that can be derived from decision theoretic foundations, such as the quantile score, WIS, and CRPS, in section \ref{sec:methods.related}.

@@

Suppose that a decision maker is tasked with determining how to allocate $K$ available units of a resource across $N$ locations.
If the decision maker is provided with a multivariate forecast $F$ where each marginal forecast distribution $F_i$ predicts resource need in a particular location, one option is to choose the resource allocation that minimizes the expected total unmet need according to the forecast.
We will give a more precise mathematical statement in section \ref{sec:methods.allocation}, but informally, the total expected unmet need according to the forecast is
\begin{align}
\sum_{i=1}^N E_{F_i}[\text{unmet need in location $i$}], \label{eqn:informal_objective}
\end{align}
where $F_i$ is the marginal forecast distribution of need in location $i$ and the unmet need in a particular location is the difference between resource need in that location and the number of resources that were allocated there.
This allocation problem has an intuitively appealing solution: allocate so that the probabilities of need exceeding allocation in various locations are as close to each other as possible.
This will lead to the allocations provided by $F$ being quantiles of the marginal distributions $F_i$ for some \emph{single} probability level $\tau$ that is shared in common for all locations.

@@

This illustrates a fundamental property of the allocation score: at its core, it measures whether the forecast accurately captures the relative magnitudes of resource need across different locations, which is precisely the information that is needed to allocate resources to those locations subject to a fixed resource constraint.
On the other hand, the allocation score is not directly sensitive to whether the forecasts in each location correctly capture the magnitude of resource need in each individual location.
%For a given probability level $\tau$, if the quantiles $Q_i(\tau)$ of the marginal forecasts are proportional to the observed needs $y_i$, then the allocation score is zero for the resource constraint level $K = \sum Q_i(\tau)$.
This stands in marked contrast to other common scoring methods for multivariate forecasts that aggregate univariate scores such as CRPS or WIS for the marginal forecasts where a 
misalignment in one coordinate is penalized regardless of alignments in other coordinates. Note that we do not claim that the allocation score is generically preferable to these other scores \textemdash rather, it provides a view of forecast performance that is specifically tuned to the context of decision making about resource allocations.

@@

\subsection{A decision theoretic development of the allocation score}
\label{sec:methods.detailed}

We give a high-level review of a general procedure for developing proper scoring rules that are tailored to specific decision making tasks in section \ref{sec:methods.detailed.decisiontheory}, and then in section \ref{sec:methods.detailed.allocation} we apply that procedure to develop the allocation score based on the task of deciding on how to allocate a fixed supply of resources across multiple locations.

\subsubsection{The decision theoretic setup for forecast evaluation}
\label{sec:methods.detailed.decisiontheory}

%In this section, we give an overview of the decision theoretic setup for developing proper scoring rules that measure the value of a forecast as an input to decision making. We keep the discussion here at a somewhat informal level; we refer the reader to [some subset of Brehmer and Gneiting; Grünwald and Dawid; Dawid; Granger and Pesaran 2000; Granger and Machina 2006; Ehm et al. 2016] for more technically precise discussion.

In the framework of decision theory, a decision corresponds to the selection of an action $x$ from some set of possible actions $\mathcal{X}$. For example, $x$ may correspond to the level of investment in a measure designed to mitigate severe disease outcomes such as hospital beds, ventilators, medication, or medical staff, with $\mathcal{X}$ being the set of all possible levels of investment that we might select. The quality of a decision to take a particular action $x$ is measured in relation to an outcome $y$ that is unknown at the time the decision is made. For example, $y$ may correspond to the number of individuals who eventually become sick and would benefit from the mitigation measure, and informally, an action $x$ is successful to the extent that it meets the realized need. In the face of uncertainty, a decision maker may use a forecast $F$ of the random variable $Y$ to help inform the selection of the action to take. We measure the value of a forecast as an input to this decision making process by the quality of the decisions that it leads to.

We can formalize the preceding discussion with the following three-step procedure for developing scoring rules for probabilistic forecasts:
\begin{enumerate}
\item Specify a \emph{loss function} $s(x, y)$ that measures the loss associated with taking action $x$ when outcome $y$ eventually occurs.
\item Given a probabilistic forecast $F$, determine the \emph{Bayes act} $x^F$ that minimizes the expected loss under the distribution $F$.
\item The \emph{scoring rule} for $F$ calculates the score as the loss incurred when the Bayes act was used: $S(F, y) = s(x^F, y)$.
\end{enumerate}
%We use the letter $s$ for the loss function to align with the literature on evaluation of forecasts of continuous outcomes, in which context we can often identify the action $x$ with a functional (i.e., a numeric summary such as a mean or a quantile) of the forecast distribution $F$. In this context, $s$ may be used as a \emph{scoring function}. \elr{Consider moving preceding sentences to a footnote or just deleting them?}
This is a general procedure that may be applied in settings where it is possible to specify a quantitative loss function. Subject to certain technical conditions, scoring rules obtained from this procedure are proper. We refer the reader to discussion in (cite paper 2 on arxiv) for a more technically precise discussion.

@@

% We review how the quantile score arises from a particular decision making problem in section \ref{sec:methods.quantileloss.quantile}, and how CRPS can be obtained by integrating across values of the parameters of that decision making problem, as well as the connection to WIS, in section \ref{sec:methods.quantileloss.crps}. These results have been thoroughly discussed in the literature [cite cite cite].

% \subsubsection{decision theoretic origins of the quantile score}
% \label{sec:methods.quantileloss.quantile}

% Suppose that a decision maker is tasked with determining the quantity $x$ of a protective measure to procure; for example, $x$ might represent the number of hospital beds or amount of medicine to purchase. Additionally, suppose that each unit of this good has cost $C$ so that the total cost of procurement is $Cx$. The variable $y$ denotes the eventual realized need for this resource, e.g. the number of patients in need of a hospital bed or the amount of medication that is needed. We assume that each unit of unmet need incurs a loss denoted by $L$, so that if the selected procurement level $x$ is less than the realized need $y$, a loss of $L(x-y)$ results. At the time that a decision maker determines the amount $x$ to procure, the demand $y$ is not yet known. We therefore define the random variable $Y$ that represents the as-yet-unknown level of demand. The forecast $F$ specifies a predictive distribution for $Y$. Here we identify $F$ with its cumulative distribution function (CDF), and $F^{-1}$ denotes the quantile function. With this formalization of the decision making task, we can proceed to develop a proper scoring rule using the procedure outlined in section \ref{sec:methods.decisiontheory}.

@@

% We have set up the problem here in terms of the cost and loss parameters $C$ and $L$, which has the benefit of an intuitive connection to the decision making context. However, to clarify the connection to the usual notation for the quantile loss, we can divide the loss function $s_Q$ by $L$ to obtain an expression in terms of only $\alpha$:
% \begin{align*}
% s_Q(x,y; \alpha) &= s_Q(x,y; C, L)/L \\
% &= (C/L)x + (x-y)_- \\
% &= (1 - \alpha)x + (x-y)_-.
% \end{align*}
% Because these loss functions are equal up to a constant of proportionality, the Bayes act is the same for both. The associated quantile scoring rule expressed in terms of $\alpha$ is
% \begin{equation}
% S_Q(F, y; \alpha) = (1 - \alpha)F^{-1}(\alpha) + \left(F^{-1}(\alpha) - y\right)_-. \label{eqn:S_Q_alpha}
% \end{equation}
% In either formulation, the key observation is that the Bayes act is the quantile of the forecast distribution $F$ at the probability level given by one minus the cost/loss ratio $C/L$.

@@

% The scoring rule $S_Q$ of Equation \eqref{eqn:S_Q_alpha} evaluates the forecast distribution $F$ only through its $\alpha$ quantile. While this is faithful to the context of the decision making problem, it may not be satisfying as a measure of the quality of the full forecast distribution. For this purpose, one option is to integrate the quantile scoring rule across different values of the probability level $\alpha$, weighting the probability levels according to a specified distribution $p$. This yields a weighted CRPS:
% \begin{align*}
% S_{CRPS}(F, y; p) &= \int S_Q(F, y; \alpha) p(\alpha) \, d\alpha.
% \end{align*}
% This weighted form of CRPS has appeared in the literature before, e.g. see \cite{gneiting2011weightedScoringRules}. The usual CRPS results from taking $p$ corresponding to a $\text{Uniform}(0,1)$ distribution, equally weighting all probability levels.

% We emphasize that because $\alpha = 1 - C/L$, the distribution $p(\alpha)$ can be interpreted as expressing incomplete knowledge about the cost/loss ratio in the decision making problem. The equal weighting used by the ordinary CRPS may be appropriate in the absence of any knowledge about the context in which forecasts will be used to support decision making, but may be inappropriate if more information is known about the cost/loss ratio for a specific decision making task.

@@

%As a concrete example, we take the resource to be a good such as ventilators or oxygen supply. An administrator is tasked with determining where to send these resources so as to meet demand among hospital patients in different facilities or states. In contrast to the decision making problem in the previous section, the administrator is not able to control the total amount of supply; rather, their task is to determine how to allocate the fixed supply to different locations. Following the structure of the discussion for the quantile score and CRPS, we first discuss the set up for a single value of the resource constraint $K$, which plays a role that is analogous to the cost/loss ratio for the quantile score, and then we discuss the possibility of integrating (or averaging) across values of $K$.

@@

In the decision making setting that we consider, an action $\mathbf{x} = (x_1, \ldots, x_N)$ is a vector specifying the amount that is allocated to each of $N$ locations. We require that each $x_i$ is non-negative and that the total allocation across all locations equals the amount of available resources, $K$: $\sum_{i=1}^N x_i = K$. The set $\mathcal{X}$ consists of all possible allocations that satisfy these constraints. The eventually realized resource need in each location is denoted by $\mathbf{y} = (y_1, \ldots, y_N)$. These levels of need are not known at the time of decision making, so we define the random vector $Y = (Y_1, \ldots, Y_N)$ where $Y_i$ represents the as-yet-unknown level of resource need in location $i$. Forecasts of need in each location are collected in $F = (F_1, \ldots, F_N)$. We assume that resource need is non-negative and the forecasts reflect that, i.e. the support of each $F_i$ is a subset of $\mathbb{R}^+$. Finally, we assume that each unit of unmet need incurs a loss denoted by $L$, so that if the selected resource level $x_i$ in location $i$ is less than the realized need $y_i$, a loss of $L(y_i - x_i)$ results.

@@

One interpretation of this result is that the Bayes act sets the allocation in each location $i$ to a quantile of the forecast distribution $F_i$ for that location. The quantile is at a probability level $\alpha = 1 - \lambda^{\star}/L$ that is the same for all locations, and is chosen such that the constraint is satisfied. According to this interpretation, we divide the available resources among the locations in such a way that the forecasted probability that need will exceed the available resources is the same in all locations.
An alternative interpretation comes from noting that for each location $i$, $\frac{\partial}{\partial x_i} \sbar_A^F(\mathbf{x}) \vert_{\mathbf{x}=\mathbf{x}^F} = \lambda^{\star}$ (see the supplement for a proof).
%$\frac{\partial}{\partial x_i} \sbar_F(\mathbf{x}) = L[F_i(x_i) - 1]$ (see the supplement for a derivation), and evaluating at the Bayes allocation we arrive at
%\begin{align}
%\frac{\partial}{\partial x_i} \sbar_F(\mathbf{x}) \vert_{\mathbf{x}=\mathbf{x}^F} &= L[F_i(x^F_i) - 1] = L[F_i(F_i^{-1}(1 - \lambda^{\star}/L)) - 1] = \lambda^{\star}.
%\end{align}
In words, at the allocation given by the Bayes act, the rate of change of the expected score as a function of the amount allocated to location $i$ is given by $\lambda^{\star}$.
This derivative is the same for all locations, so the optimal allocation divides the available resources across all locations in such a way that according to $F$, the expected benefit of 1 additional unit of resources is the same in all locations.

@@

Case study heading into the Omicron wave. Some more detailed discussion of implications of bad forecasts for specific decision making purposes -- take a "deep dive" into one or two example states like FL.
